|

### 2023-02-08 paper.tex
Wed Feb  8 21:00:07 CET 2023: update-remote.sh

Denoising Autoencoder Genetic Programming (DAE-GP) is a novel variation of an genetic programming based Estimation of Distribution Algorithm (EDA-GP) that uses a denoising autoencoders long short-term memory network (DAE-LSTMs) as a probabilistic model to sample new candidate solutions {[}\protect\hyperlink{ref-dae-gp_2020_rtree}{33}{]}.

DAE-LSTMs are artificial neural networks that can be trained in an unsupervised learning environment to minimize a reconstruction error for encoding input data into a compressed representation and subsequently decoding the compressed representation back to the input dimension.
In DAE-GP, DAE-LSTMs are trained with a subset of high-fitness solutions selected from a parent population with the aim to capture their promising qualities.
The resulting model is then used to sample new offspring solutions by propagating partially mutated solutions from the parent population through the DAE-LSTMS {[}\protect\hyperlink{ref-dae-gp_2020_rtree}{33}{]}.
In previous work DAE-GP has been shown to outperform GP for both a generalized version of the royal tree problem {[}\protect\hyperlink{ref-dae-gp_2020_rtree}{33}{]} as well as for a real-world symbolic regression problem {[}\protect\hyperlink{ref-dae-gp_2022_symreg}{32}{]}.

The DAE-GP algorithm first described by {[}\protect\hyperlink{ref-dae-gp_2020_rtree}{33}{]} trains a DAE-LSTMs for each generation \(g\) of the search from scratch.
{[}\protect\hyperlink{ref-dae-gp_2022_symreg}{32}{]} and {[}\protect\hyperlink{ref-daegp_explore_exploit}{31}{]} suggests the incorporation of a pre-training strategy into the evolutionary search as a possible way of improving the overall performance of the DAE-GP algorithm.
The key idea is to pre-train an initial DAE-LSTM on a large population of candidate solutions and to use the pre-trained parameters of this initial model in each generation of the search as starting parameters for the current generation DAE-LSTM.
This thesis studies the influence of using pre-trained DAE-LSTMs in DAE-GP for symbolic regression, especially looking at the influence on overall quality of solutions found by DAE-GP and effects on run-time.
The aim of this study is to answer the question if a pre-training strategy can be used to improve DAE-GP and to study both positive and negative effects on overall performance.

@@

DAE-GP is an EDA-GP algorithm that uses DAE-LSTM networks as a probabilistic model to sample new offspring solutions {[}\protect\hyperlink{ref-dae-gp_2020_rtree}{33}{]}.

@@

DAE was first introduced by {[}\protect\hyperlink{ref-dae_orig2008}{27}{]} as an improved AE with the ability to learn new representations of data that is especially robust to partially corrupted input data. DAE modifies the AE by using partially corrupted input data for the AE and training it to reconstruct the uncorrupted, original version of the input data.

@@

DAE-GP builds upon the concept of using DAEs in EDAs described by {[}\protect\hyperlink{ref-harmless_overfitting_eda}{21}{]} and transfers the concept to the domain of GP.
The mutation and crossover operators of standard GP are replaced by sampling new solutions from a probabilistic model that is build by training a DAEs long short-term memory (LSTM) network on a subset of high fitness solutions from the current population {[}\protect\hyperlink{ref-dae-gp_2020_rtree}{33}{]}.

LSTMs are a variant of artificial neural networks first introduced by {[}\protect\hyperlink{ref-lstm_orig}{11}{]} that can store learned information over en extended time period while avoiding the problem of vanishing and/or exploding gradients. Since DAE-GP encodes candidate solutions as linear strings in prefix notation, the DAE in DAE-GP uses LSTMs for both encoding and decoding where the total amount of time steps \(T\) is equal to sum of the length of the input solution and the output solution {[}\protect\hyperlink{ref-dae-gp_2020_rtree}{33}{]}.

@@

The pre-training strategy used in this thesis is applied to the DAE-LSTM model \(M_g\) that are used in each DAE-GP generation \(g\) for \(g\in{1,...,g_{max}}\).
Instead of initializing and optimizing each model from scratch as done in previous work (e.g. {[}\protect\hyperlink{ref-dae-gp_2020_rtree}{33}{]} {[}\protect\hyperlink{ref-dae-gp_2022_symreg}{32}{]}), a separate DAE-LSTM network \(\hat{M}\) will be trained on a large population of randomly initialized solutions.\\
The trainable parameters \(\theta_{\hat{M}}\) obtained after finishing the training procedure of \(\hat{M}\) are then used as the starting parameters for each following DAE-LSTM model \(M_g\) for \(g\in{1,...,g_{max}}\).

@@

The DAE-GP algorithm, first described by {[}\protect\hyperlink{ref-dae-gp_2020_rtree}{33}{]}, is summarized as a flowchart in figure \ref{fig:daegp-flowchart}.
After setting the generation counter \(g\) to \(0\) and creating the initial population of solutions \(P_0\), DAE-GP enters the main loop that checks if a termination criteria is satisfied (i.e.~maximum number of generations or fitness evaluations is reached).
For each generation \(g\) the fitness of all new individuals in \(P_g\) is evaluated. During model building, DAE-GP selects a subset \(X_g\) of the current population that is used as training data for the DAE-LSTM model \(M_g\) which is trained to learn the properties of \(X_g\).
The trained model \(M_g\) is then used for model sampling, partially mutated solutions from \(P_g\) are propagated through \(M_g\) to create a new population \(P_{g+1}\).
This process is repeated until the termination criteria is met and DAE-GP returns the highest fitness solutions.

@@

One difficulty in the implementation of a pre-training strategy into DAE-GP has been the determination of the number of hidden neurons in side the DAE-LSTMs hidden layers. The original description of DAE-GP {[}\protect\hyperlink{ref-dae-gp_2020_rtree}{33}{]} uses a strategy where the number of hidden neurons for each hidden layer is dynamically set per generation to the maximum individual size inside the current population.
For a pre-training implementation, this strategy can not be easily adapted since it leads to a changing number of neurons at each generation resulting in different dimensions of the DAE-LSTM. To allow the sharing of the state of the pre-trained parameters \(\theta_{\hat{M}}\) from the pre-trained model \(\hat{M}\) to each \(M_g\) for \(g\in{1,...,g_{max}}\) the number of hidden neurons per hidden layer for all DAE-GP runs, pre-trained as well as regular, is statically set to a predefined value.

@@

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.5116}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.4884}}@{}}
\caption{\label{tab:baseParams-DAEGP}DAE-LSTM - Hyperparameter}\tabularnewline
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Hyperparameter
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Value
\end{minipage} \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Hyperparameter
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Value
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
Stopping Condition & Convergence {[}\protect\hyperlink{ref-daegp_explore_exploit}{31}{]} \\
Reconstruction Error Metric & Multiclass Cross Entropy \\
Gradient Descent Algorithm & Adam Optimization {[}\protect\hyperlink{ref-Kingma2014AdamAM}{13}{]} \\
Learning Rate & 0.001 \\
Batch Size & 0.1 \\
Sampling Steps & 1 \\
Hidden Layer & 2 \\
Hidden Neurons & 150 \\
Corruption Operator & Levenshtein Edit {[}\protect\hyperlink{ref-daegp_explore_exploit}{31}{]} \\
Denoising Edit Probability (Model Building) & 0.05 \\
Denoising Edit Probability (Model Sampling) & 0.95 \\
Pre-Training Stopping Condition & Early Stopping \\
\end{longtable}

The framework that was used to run all experiments of this thesis was provided by the supervisor of this thesis David Wittenberg and uses the python programming language {[}\protect\hyperlink{ref-python3}{26}{]} in conjunction with the baseline libraries \texttt{Keras} {[}\protect\hyperlink{ref-chollet2015keras}{2}{]} for deep neural networks and \texttt{deap} {[}\protect\hyperlink{ref-DEAP_JMLR2012}{6}{]} for evolutionary computation. For data visualizations I used the python package \texttt{matplotlib} {[}\protect\hyperlink{ref-matplotlib}{12}{]}, statistical analysis and evaluations were conducted using primarily the data science libraries \texttt{scipy} {[}\protect\hyperlink{ref-scipy}{28}{]}, \texttt{numpy} {[}\protect\hyperlink{ref-numpy}{8}{]}, \texttt{pandas} {[}\protect\hyperlink{ref-pandas}{24}{]}, and \texttt{IPython} {[}\protect\hyperlink{ref-ipython}{20}{]}.

@@

The objective of the airfoil problem is to find a function that accurately predicts the output variable \(y\) by taking in a subset of the input variables \(x1,x2,x3,x4,x5\). The function set used by all DAE-GP variations for symbolic regression is summarized in tables \ref{tab:baseParams-GP} and , the terminal set consists of the the input variables \(x1,x2,x3,x4,x5\) and the ephemeral integer constants listed in table \ref{tab:baseParams-GP} {[}\protect\hyperlink{ref-dae-gp_2022_symreg}{32}{]}.

@@

In general I expect that with a growing number of trainable parameters (either by adding more hidden layers or more hidden neurons per layer) the DAE-LSTM will be more prone to overfit to the training population \(P_{train}\) resulting in a small reconstruction error for the training population and a large one for the validation population \(P_{test}\). The reason for this effect is that a larger network trained over an extended period of time (without the use of strategies like early stopping), has much more potential to learn noise from the training dataset than a smaller network, which is more likely to result in worsening performance for previously unseen cases {[}\protect\hyperlink{ref-weigend1994overfitting}{29}{]}.

@@

Since previous work by {[}\protect\hyperlink{ref-dae-gp_2022_symreg}{32}{]} demonstrated that DAE-GP, for a given number of 10.000 fitness evaluations, is able to produce solutions that are much smaller in tree size than regular GP, I also studied the influence of pre-training on the size of solutions found by DAE-GP. Figure \ref{fig:airfoil-size-combined-30gens} shows the median of both the average size of individuals inside the population as well as the size of the currently best performing solution inside the population. It can be observed that both DAE-GP variants strongly reduce the average solutions size already in the first generations and that the average solution size inside the population constantly stays at this very low level for the rest of the search. The median average size over all generations is 1.536 for DAE-GP and 1.536 for pre-trained DAE-GP. Regarding the size of the best solution per generation, the results show that the pre-trained version of algorithm did produce solutions with the same median size as traditional DAE-GP. The median size of the best solutions found after terminating the search is the same for both algorithms (DAE-GP: 5, Pre-Trained DAE-GP: 5)

@@

To ensure a fair comparison between pre-trained and regular DAE-GP, the prior experiments used a static number of hidden neurons for each hidden Layer inside the DAE-LSTM networks. For comparison, figure \ref{fig:airfoil-maxIndSize-params} shows the total number of trainable parameters for each generation's DAE-LSTM of another experiment where regular DAE-GP was tested with dynamic adjustment of the number of hidden neurons as described in {[}\protect\hyperlink{ref-dae-gp_2020_rtree}{33}{]} and {[}\protect\hyperlink{ref-dae-gp_2022_symreg}{32}{]}. Again, the experiments was based on \(10\) individual runs and otherwise used the same hyperparameters as listed in tables \ref{tab:baseParams-GP} and \ref{tab:baseParams-DAEGP}. It clearly shows that DAE-GP can strongly reduce the number of trainable parameters by adjusting the number of hidden neurons per hidden layers to the current generations population which largely reduces the amount of training necessary and likely reduce the run-time of DAE-GP.

@@

Since previous research on DAE-GP used a single hidden layers for the airfoil dataset {[}\protect\hyperlink{ref-dae-gp_2022_symreg}{32}{]}, I repeated the experiment with a single hidden layer. Based on the findings of {[}\protect\hyperlink{ref-pmlr-v5-erhan09a}{5}{]}, I expected that pre-training DAE-GP would lead to a decrease in overall performance. Figures \ref{fig:airfoil-fitness-perGen} and \ref{fig:airfoil-finalFitness} shows the mean and median best fitness per generation as well as the distribution of final fitness scores after 30 generations. The results are, again, based on \(10\) individual runs per algorithm. As expected the results show, that for a small DAE-LSTM size pre-training negatively impacts the fitness of solution found by DAE-GP. The median final RMSE for the test set of pre-trained DAE-GP increases by a factor of 1.287 to 16.213 compared to regular DAE-GP with 12.594.

@@

\begin{longtable}[]{@{}lrrl@{}}
\caption{\label{tab:full-run-realWorldSymReg-problems}Real World Symbolic Regression Benchmark Problems}\tabularnewline
\toprule\noalign{}
Problem & Observations & Features & Source \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
Problem & Observations & Features & Source \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
Airfoil & 1503 & 5 & {[}\protect\hyperlink{ref-AIRFOIL_DATASET}{1}{]} \\
Boston\_Housing & 506 & 13 & {[}\protect\hyperlink{ref-BOSTON_HOUSING_DATASET}{9}{]} \\
Energy\_Cooling & 768 & 8 & {[}\protect\hyperlink{ref-ENERGY_DATASET}{25}{]} \\
Concrete & 1030 & 8 & {[}\protect\hyperlink{ref-CONCRETE_DATASET}{34}{]} \\
\end{longtable}

To test for statistical differences of the results achieved by either pre-trained or regular DAE-GP, I selected the nonparametric Mann-Whitney-U Test ({[}\protect\hyperlink{ref-mwu1}{30}{]}, {[}\protect\hyperlink{ref-mwu2}{16}{]}) to test for differences in the distribution underlying the samples sourced from my experiments. Additionally, I included the effect strength of pre-training as measured by Cliffs Delta {[}\protect\hyperlink{ref-Cliff1993DominanceSO}{3}{]} into the statistical analysis. Better median values (e.g.~higher fitness) are printed bold and p-values are marked with asterisk (1,2,3) for \(\alpha\) levels (\(0.1\), \(0.05\), \(0.01\)).

Note that the results for the real world symbolic regression problems inside this chapter, except for the airfoil problem (see subsection \protect\hyperlink{influence-on-run-time}{Influence on Run-Time}), have not been analyzed for their total run-time. This is due to the fact, that experiments were run on different computers architectures which resulted in strongly biased results.

@@

The results for the median best fitness over the evolutionary search are visualized in figure \ref{fig:symReg-Fitness}. Figure \ref{fig:symReg-Fitness-BoxPlots} shows the distribution of final fitness values on both the test and training set as box plots.

@@

@@

Table \ref{tab:full-run-realWorldSymReg-fitness} summarizes the results for the median best fitness achieved after 30 generations and shows the results of computing the statistical tests.

@@

Another disadvantage of using pre-training in DAE-GP is that, in my experiments, it required DAE-LSTM networks to use two hidden layers (see subsection \protect\hyperlink{reduced-number-of-hidden-layers}{Reduced Number of Hidden Layers}). For a single hidden layer, pre-training did show a negative impact on the median final fitness of solutions found as it was suggested by {[}\protect\hyperlink{ref-pmlr-v9-erhan10a}{4}{]}. Since current research on DAE-GP for real-world symbolic regression relies on a single hidden layer (see {[}\protect\hyperlink{ref-dae-gp_2022_symreg}{32}{]}), pre-training does come with additional computational expenses for using deeper DAE-LSTMs.

@@

\leavevmode\vadjust pre{\hypertarget{ref-dae_orig2008}{}}%
\CSLLeftMargin{{[}27{]} }%
\CSLRightInline{Pascal Vincent, Hugo Larochelle, Y. Bengio, and Pierre-Antoine Manzagol. 2008. Extracting and composing robust features with denoising autoencoders. In \emph{Proceedings of the 25th international conference on machine learning} (ICML '08), Association for Computing Machinery, New York, NY, USA, 1096--1103. DOI:https://doi.org/\href{https://doi.org/10.1145/1390156.1390294}{10.1145/1390156.1390294}}

\leavevmode\vadjust pre{\hypertarget{ref-scipy}{}}%
\CSLLeftMargin{{[}28{]} }%
\CSLRightInline{Pauli Virtanen, Ralf Gommers, Travis E. Oliphant, Matt Haberland, Tyler Reddy, David Cournapeau, Evgeni Burovski, Pearu Peterson, Warren Weckesser, Jonathan Bright, Stéfan J. van der Walt, Matthew Brett, Joshua Wilson, K. Jarrod Millman, Nikolay Mayorov, Andrew R. J. Nelson, Eric Jones, Robert Kern, Eric Larson, C J Carey, İlhan Polat, Yu Feng, Eric W. Moore, Jake VanderPlas, Denis Laxalde, Josef Perktold, Robert Cimrman, Ian Henriksen, E. A. Quintero, Charles R. Harris, Anne M. Archibald, Antônio H. Ribeiro, Fabian Pedregosa, Paul van Mulbregt, and SciPy 1.0 Contributors. 2020. {{SciPy} 1.0: Fundamental Algorithms for Scientific Computing in Python}. \emph{Nature Methods} 17, (2020), 261--272. DOI:https://doi.org/\href{https://doi.org/10.1038/s41592-019-0686-2}{10.1038/s41592-019-0686-2}}

>>>>

Denoising Autoencoder Genetic Programming (DAE-GP) is a novel variation of an genetic programming based Estimation of Distribution Algorithm (EDA-GP) that uses a denoising autoencoders long short-term memory network (DAE-LSTMs) as a probabilistic model to sample new candidate solutions {[}\protect\hyperlink{ref-dae-gp_2020_rtree}{34}{]}.

DAE-LSTMs are artificial neural networks that can be trained in an unsupervised learning environment to minimize a reconstruction error for encoding input data into a compressed representation and subsequently decoding the compressed representation back to the input dimension.
In DAE-GP, DAE-LSTMs are trained with a subset of high-fitness solutions selected from a parent population with the aim to capture their promising qualities.
The resulting model is then used to sample new offspring solutions by propagating partially mutated solutions from the parent population through the DAE-LSTMS {[}\protect\hyperlink{ref-dae-gp_2020_rtree}{34}{]}.
In previous work DAE-GP has been shown to outperform GP for both a generalized version of the royal tree problem {[}\protect\hyperlink{ref-dae-gp_2020_rtree}{34}{]} as well as for a real-world symbolic regression problem {[}\protect\hyperlink{ref-dae-gp_2022_symreg}{33}{]}.

The DAE-GP algorithm first described by {[}\protect\hyperlink{ref-dae-gp_2020_rtree}{34}{]} trains a DAE-LSTMs for each generation \(g\) of the search from scratch.
{[}\protect\hyperlink{ref-dae-gp_2022_symreg}{33}{]} and {[}\protect\hyperlink{ref-daegp_explore_exploit}{32}{]} suggests the incorporation of a pre-training strategy into the evolutionary search as a possible way of improving the overall performance of the DAE-GP algorithm.
The key idea is to pre-train an initial DAE-LSTM on a large population of candidate solutions and to use the pre-trained parameters of this initial model in each generation of the search as starting parameters for the current generation DAE-LSTM.
This thesis studies the influence of using pre-trained DAE-LSTMs in DAE-GP for symbolic regression, especially looking at the influence on overall quality of solutions found by DAE-GP and effects on run-time.
The aim of this study is to answer the question if a pre-training strategy can be used to improve DAE-GP and to study both positive and negative effects on overall performance.

@@

DAE-GP is an EDA-GP algorithm that uses DAE-LSTM networks as a probabilistic model to sample new offspring solutions {[}\protect\hyperlink{ref-dae-gp_2020_rtree}{34}{]}.

@@

DAE was first introduced by {[}\protect\hyperlink{ref-dae_orig2008}{28}{]} as an improved AE with the ability to learn new representations of data that is especially robust to partially corrupted input data. DAE modifies the AE by using partially corrupted input data for the AE and training it to reconstruct the uncorrupted, original version of the input data.

@@

DAE-GP builds upon the concept of using DAEs in EDAs described by {[}\protect\hyperlink{ref-harmless_overfitting_eda}{21}{]} and transfers the concept to the domain of GP.
The mutation and crossover operators of standard GP are replaced by sampling new solutions from a probabilistic model that is build by training a DAEs long short-term memory (LSTM) network on a subset of high fitness solutions from the current population {[}\protect\hyperlink{ref-dae-gp_2020_rtree}{34}{]}.

LSTMs are a variant of artificial neural networks first introduced by {[}\protect\hyperlink{ref-lstm_orig}{11}{]} that can store learned information over en extended time period while avoiding the problem of vanishing and/or exploding gradients. Since DAE-GP encodes candidate solutions as linear strings in prefix notation, the DAE in DAE-GP uses LSTMs for both encoding and decoding where the total amount of time steps \(T\) is equal to sum of the length of the input solution and the output solution {[}\protect\hyperlink{ref-dae-gp_2020_rtree}{34}{]}.

@@

The pre-training strategy used in this thesis is applied to the DAE-LSTM model \(M_g\) that are used in each DAE-GP generation \(g\) for \(g\in{1,...,g_{max}}\).
Instead of initializing and optimizing each model from scratch as done in previous work (e.g. {[}\protect\hyperlink{ref-dae-gp_2020_rtree}{34}{]} {[}\protect\hyperlink{ref-dae-gp_2022_symreg}{33}{]}), a separate DAE-LSTM network \(\hat{M}\) will be trained on a large population of randomly initialized solutions.\\
The trainable parameters \(\theta_{\hat{M}}\) obtained after finishing the training procedure of \(\hat{M}\) are then used as the starting parameters for each following DAE-LSTM model \(M_g\) for \(g\in{1,...,g_{max}}\).

@@

The DAE-GP algorithm, first described by {[}\protect\hyperlink{ref-dae-gp_2020_rtree}{34}{]}, is summarized as a flowchart in figure \ref{fig:daegp-flowchart}.
After setting the generation counter \(g\) to \(0\) and creating the initial population of solutions \(P_0\), DAE-GP enters the main loop that checks if a termination criteria is satisfied (i.e.~maximum number of generations or fitness evaluations is reached).
For each generation \(g\) the fitness of all new individuals in \(P_g\) is evaluated. During model building, DAE-GP selects a subset \(X_g\) of the current population that is used as training data for the DAE-LSTM model \(M_g\) which is trained to learn the properties of \(X_g\).
The trained model \(M_g\) is then used for model sampling, partially mutated solutions from \(P_g\) are propagated through \(M_g\) to create a new population \(P_{g+1}\).
This process is repeated until the termination criteria is met and DAE-GP returns the highest fitness solutions.

@@

One difficulty in the implementation of a pre-training strategy into DAE-GP has been the determination of the number of hidden neurons in side the DAE-LSTMs hidden layers. The original description of DAE-GP {[}\protect\hyperlink{ref-dae-gp_2020_rtree}{34}{]} uses a strategy where the number of hidden neurons for each hidden layer is dynamically set per generation to the maximum individual size inside the current population.
For a pre-training implementation, this strategy can not be easily adapted since it leads to a changing number of neurons at each generation resulting in different dimensions of the DAE-LSTM. To allow the sharing of the state of the pre-trained parameters \(\theta_{\hat{M}}\) from the pre-trained model \(\hat{M}\) to each \(M_g\) for \(g\in{1,...,g_{max}}\) the number of hidden neurons per hidden layer for all DAE-GP runs, pre-trained as well as regular, is statically set to a predefined value.

@@

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.5116}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.4884}}@{}}
\caption{\label{tab:baseParams-DAEGP}DAE-LSTM - Hyperparameter}\tabularnewline
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Hyperparameter
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Value
\end{minipage} \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Hyperparameter
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Value
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
Stopping Condition & Convergence {[}\protect\hyperlink{ref-daegp_explore_exploit}{32}{]} \\
Reconstruction Error Metric & Multiclass Cross Entropy \\
Gradient Descent Algorithm & Adam Optimization {[}\protect\hyperlink{ref-Kingma2014AdamAM}{13}{]} \\
Learning Rate & 0.001 \\
Batch Size & 0.1 \\
Sampling Steps & 1 \\
Hidden Layer & 2 \\
Hidden Neurons & 150 \\
Corruption Operator & Levenshtein Edit {[}\protect\hyperlink{ref-daegp_explore_exploit}{32}{]} \\
Denoising Edit Probability (Model Building) & 0.05 \\
Denoising Edit Probability (Model Sampling) & 0.95 \\
Pre-Training Stopping Condition & Early Stopping \\
\end{longtable}

The framework that was used to run all experiments of this thesis was provided by the supervisor of this thesis David Wittenberg and uses the python programming language {[}\protect\hyperlink{ref-python3}{26}{]} in conjunction with the baseline libraries \texttt{Keras} {[}\protect\hyperlink{ref-chollet2015keras}{2}{]} for deep neural networks and \texttt{deap} {[}\protect\hyperlink{ref-DEAP_JMLR2012}{6}{]} for evolutionary computation. For data visualizations I used the python package \texttt{matplotlib} {[}\protect\hyperlink{ref-matplotlib}{12}{]}, statistical analysis and evaluations were conducted using primarily the data science libraries \texttt{scipy} {[}\protect\hyperlink{ref-scipy}{29}{]}, \texttt{numpy} {[}\protect\hyperlink{ref-numpy}{8}{]}, \texttt{pandas} {[}\protect\hyperlink{ref-pandas}{24}{]}, and \texttt{IPython} {[}\protect\hyperlink{ref-ipython}{20}{]}.

@@

The objective of the airfoil problem is to find a function that accurately predicts the output variable \(y\) by taking in a subset of the input variables \(x1,x2,x3,x4,x5\). The function set used by all DAE-GP variations for symbolic regression is summarized in tables \ref{tab:baseParams-GP} and , the terminal set consists of the the input variables \(x1,x2,x3,x4,x5\) and the ephemeral integer constants listed in table \ref{tab:baseParams-GP} {[}\protect\hyperlink{ref-dae-gp_2022_symreg}{33}{]}.

@@

In general I expect that with a growing number of trainable parameters (either by adding more hidden layers or more hidden neurons per layer) the DAE-LSTM will be more prone to overfit to the training population \(P_{train}\) resulting in a small reconstruction error for the training population and a large one for the validation population \(P_{test}\). The reason for this effect is that a larger network trained over an extended period of time (without the use of strategies like early stopping), has much more potential to learn noise from the training dataset than a smaller network, which is more likely to result in worsening performance for previously unseen cases {[}\protect\hyperlink{ref-weigend1994overfitting}{30}{]}.

@@

Since previous work by {[}\protect\hyperlink{ref-dae-gp_2022_symreg}{33}{]} demonstrated that DAE-GP, for a given number of 10.000 fitness evaluations, is able to produce solutions that are much smaller in tree size than regular GP, I also studied the influence of pre-training on the size of solutions found by DAE-GP. Figure \ref{fig:airfoil-size-combined-30gens} shows the median of both the average size of individuals inside the population as well as the size of the currently best performing solution inside the population. It can be observed that both DAE-GP variants strongly reduce the average solutions size already in the first generations and that the average solution size inside the population constantly stays at this very low level for the rest of the search. The median average size over all generations is 1.536 for DAE-GP and 1.536 for pre-trained DAE-GP. Regarding the size of the best solution per generation, the results show that the pre-trained version of algorithm did produce solutions with the same median size as traditional DAE-GP. The median size of the best solutions found after terminating the search is the same for both algorithms (DAE-GP: 5, Pre-Trained DAE-GP: 5)

@@

To ensure a fair comparison between pre-trained and regular DAE-GP, the prior experiments used a static number of hidden neurons for each hidden Layer inside the DAE-LSTM networks. For comparison, figure \ref{fig:airfoil-maxIndSize-params} shows the total number of trainable parameters for each generation's DAE-LSTM of another experiment where regular DAE-GP was tested with dynamic adjustment of the number of hidden neurons as described in {[}\protect\hyperlink{ref-dae-gp_2020_rtree}{34}{]} and {[}\protect\hyperlink{ref-dae-gp_2022_symreg}{33}{]}. Again, the experiments was based on \(10\) individual runs and otherwise used the same hyperparameters as listed in tables \ref{tab:baseParams-GP} and \ref{tab:baseParams-DAEGP}. It clearly shows that DAE-GP can strongly reduce the number of trainable parameters by adjusting the number of hidden neurons per hidden layers to the current generations population which largely reduces the amount of training necessary and likely reduce the run-time of DAE-GP.

@@

Since previous research on DAE-GP used a single hidden layers for the airfoil dataset {[}\protect\hyperlink{ref-dae-gp_2022_symreg}{33}{]}, I repeated the experiment with a single hidden layer. Based on the findings of {[}\protect\hyperlink{ref-pmlr-v5-erhan09a}{5}{]}, I expected that pre-training DAE-GP would lead to a decrease in overall performance. Figures \ref{fig:airfoil-fitness-perGen} and \ref{fig:airfoil-finalFitness} shows the mean and median best fitness per generation as well as the distribution of final fitness scores after 30 generations. The results are, again, based on \(10\) individual runs per algorithm. As expected the results show, that for a small DAE-LSTM size pre-training negatively impacts the fitness of solution found by DAE-GP. The median final RMSE for the test set of pre-trained DAE-GP increases by a factor of 1.287 to 16.213 compared to regular DAE-GP with 12.594.

@@

\begin{longtable}[]{@{}lrrl@{}}
\caption{\label{tab:full-run-realWorldSymReg-problems}Real World Symbolic Regression Benchmark Problems}\tabularnewline
\toprule\noalign{}
Problem & Observations & Features & Source \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
Problem & Observations & Features & Source \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
Airfoil & 1503 & 5 & {[}\protect\hyperlink{ref-AIRFOIL_DATASET}{1}{]} \\
Boston\_Housing & 506 & 13 & {[}\protect\hyperlink{ref-BOSTON_HOUSING_DATASET}{9}{]} \\
Energy\_Cooling & 768 & 8 & {[}\protect\hyperlink{ref-ENERGY_DATASET}{25}{]} \\
Concrete & 1030 & 8 & {[}\protect\hyperlink{ref-CONCRETE_DATASET}{35}{]} \\
\end{longtable}

To test for statistical differences of the results achieved by either pre-trained or regular DAE-GP, I selected the nonparametric Mann-Whitney-U Test ({[}\protect\hyperlink{ref-mwu1}{31}{]}, {[}\protect\hyperlink{ref-mwu2}{16}{]}) to test for differences in the distribution underlying the samples sourced from my experiments. Additionally, I included the effect strength of pre-training as measured by Cliffs Delta {[}\protect\hyperlink{ref-Cliff1993DominanceSO}{3}{]} into the statistical analysis. Better median values (e.g.~higher fitness) are printed bold and p-values are marked with asterisk (1,2,3) for \(\alpha\) levels (\(0.1\), \(0.05\), \(0.01\)).

Note that the results for the real world symbolic regression problems inside this chapter, except for the airfoil problem (see subsection \protect\hyperlink{influence-on-run-time}{Influence on Run-Time}), have not been analyzed for their total run-time. This is due to the fact, that experiments were run on different computers architectures which would result in biased results.

@@

@@

The results for the median best fitness over the evolutionary search are visualized in figure \ref{fig:symReg-Fitness}. In general, pre-training and regular DAE-GP both are able to improve on the best fitness found over the evolutionary search. Looking at figure \ref{fig:symReg-Fitness}, it appears that no significant differences for the best fitness between both algorithms exist. Regular DAE-GP achieves a slightly lower RMSE on the test set for Airfoil, Boston Housing and the Energy Cooling dataset while pre-trained DAE-GP is slightly better for the Concrete dataset.

Table \ref{tab:full-run-realWorldSymReg-fitness} summarizes the results for the median best fitness achieved after 30 generations and shows the results of computing the statistical tests. As expected after reviewing figure \ref{fig:symReg-Fitness}, none of the differences in median best fitness are statistically significant and pre-training has only a very small, negligible effect strength for Cliff's Delta ranging between -0.04 and 0.17 {[}\protect\hyperlink{ref-Vargha2000ACA}{27}{]}.

Figure \ref{fig:symReg-Fitness-BoxPlots} shows the distribution of final fitness values on both the test and training set as box plots. It shows that depending on the symbolic regression problem, the dispersion of fitness values is different for both DAE-GP variants. For example, a comparativley large span is visible for the airfoil problem while fitness scores for Energy Cooling are much more concentrated. The distribution are nontheless similar for both algorithms which does not suggest any meaningful differences between using pre-training and the distribution of final fitness scores.

@@

@@

Another disadvantage of using pre-training in DAE-GP is that, in my experiments, it required DAE-LSTM networks to use two hidden layers (see subsection \protect\hyperlink{reduced-number-of-hidden-layers}{Reduced Number of Hidden Layers}). For a single hidden layer, pre-training did show a negative impact on the median final fitness of solutions found as it was suggested by {[}\protect\hyperlink{ref-pmlr-v9-erhan10a}{4}{]}. Since current research on DAE-GP for real-world symbolic regression relies on a single hidden layer (see {[}\protect\hyperlink{ref-dae-gp_2022_symreg}{33}{]}), pre-training does come with additional computational expenses for using deeper DAE-LSTMs.

@@

\leavevmode\vadjust pre{\hypertarget{ref-Vargha2000ACA}{}}%
\CSLLeftMargin{{[}27{]} }%
\CSLRightInline{András Vargha and Harold D Delaney. 2000. A critique and improvement of the CL common language effect size statistics of McGraw and wong. \emph{Journal of Educational and Behavioral Statistics} 25, (2000), 101--132.}

\leavevmode\vadjust pre{\hypertarget{ref-dae_orig2008}{}}%
\CSLLeftMargin{{[}28{]} }%
\CSLRightInline{Pascal Vincent, Hugo Larochelle, Y. Bengio, and Pierre-Antoine Manzagol. 2008. Extracting and composing robust features with denoising autoencoders. In \emph{Proceedings of the 25th international conference on machine learning} (ICML '08), Association for Computing Machinery, New York, NY, USA, 1096--1103. DOI:https://doi.org/\href{https://doi.org/10.1145/1390156.1390294}{10.1145/1390156.1390294}}
